---
title: "STATA_group_proj"
author: "Rithu Uppalapati"
date: "11/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### STATA
```{r}
#devtools::install_github("Hemken/Statamarkdown")
library(Statamarkdown)
```

```{stata}
import delimited "/Users/rithuuppalapati/Desktop/nhanes.csv", encoding(ISO-8859-1)
```
#### Initial Data Balance:
This will allow us to determine, which covariates will be useful when calculating the propensity score. 
* To-Do 
  + Create Confidence Intervals for all of the variables in order to know which to choose in our final model
  + create a standardized balance table and compare the means 
```{stata}
logit heart_attack relative_heart_attack gender age race edu annual_income 
weight bmi diabete smoke_life phy_vigorous phy_moderate blood_press 
blood_press2 hyper_med hbq_med high_chol meadial_access cover_hc health_diet
 year_smoke year_hyper
 
ttest diabete, by(heart_attack)
```

#### I. Propensity score estimation:

```{stata}
pscore diabete relative_heart_attack gender age race edu annual_income 
weight bmi smoke_life phy_vigorous phy_moderate blood_press blood_press2 
hyper_med hbq_med high_chol meadial_access cover_hc health_diet year_smoke 
year_hyper, pscore(pc_pscore) blockid(pc_block) detail

psgraph, treated(diabete)pscore(pc_pscore)
```

#### II. Propensity score matching/weighting:
*To-Do
  + create graphics 
  + for matching, I will create a line graph and juxtapose the untreated vs. treated 
  + for weighting, I will create two density graphs to compare 

##### Matching 

```{stata}
* Match pscore
qui psmatch2 diabete, kernel outcome(heart_attack) pscore(pc_pscore)
```

##### Inverse Propensity Score Weighting 
```{stata}
*inverse-weighting 
 qui dr heart_attack diabete relative_heart_attack gender age race edu 
 annual_income weight bmi smoke_life phy_vigorous phy_moderate blood_press
 blood_press2 hyper_med hbq_med high_chol meadial_access cover_hc health_diet 
 year_smoke year_hyper, genvars
 egen sumofweights = total(iptwt)
 gen norm_weights = iptwt/sumofweights
```

#### III. Balance Checking:
* To-Do 
  + I will also a linear regression estimation using the teffects ra function 
  
#### Matched Data Balance 
```{stata}
psmatch2 diabete relative_heart_attack gender age race edu annual_income 
 weight bmi smoke_life phy_vigorous phy_moderate blood_press blood_press2 
 hyper_med hbq_med high_chol meadial_access cover_hc health_diet year_smoke 
 year_hyper, out(heart_attack)
 
 ***regression on matched data 
 reg heart_attack relative_heart_attack gender age race edu 
 annual_income weight bmi smoke_life phy_vigorous phy_moderate 
 blood_press blood_press2 hyper_med hbq_med high_chol meadial_access 
 cover_hc health_diet year_smoke year_hyper diabete [fweight = _weight]
 
 ***effects 
  teffects ipw (heart_attack) (diabete relative_heart_attack gender age race 
  edu annual_income weight bmi smoke_life phy_vigorous phy_moderate blood_press
  blood_press2 hyper_med hbq_med high_chol meadial_access cover_hc health_diet 
  year_smoke year_hyper)
```

#### Inverse Weighting Balance

```{stata}
**inverse-weighting 
 pbalchk diabete heart_attack diabete relative_heart_attack gender age race edu
 annual_income weight bmi smoke_life phy_vigorous phy_moderate blood_press 
 blood_press2 hyper_med hbq_med high_chol meadial_access cover_hc health_diet 
 year_smoke year_hyper, wt(norm_weights)
 
 *** regression
 teffects ipwra (heart_attack relative_heart_attack gender age race edu 
 annual_income weight bmi smoke_life phy_vigorous phy_moderate blood_press 
 blood_press2 hyper_med hbq_med high_chol meadial_access cover_hc health_diet 
 year_smoke year_hyper) (diabete relative_heart_attack gender age race edu 
 annual_income weight bmi smoke_life phy_vigorous phy_moderate blood_press 
 blood_press2 hyper_med hbq_med high_chol meadial_access cover_hc health_diet 
 year_smoke year_hyper)
```

#### IV. Comparison between pre_match data and matched data.
* To-Do
  + Create Graphs comparing pre-match data and match data 
  + Create graphs comparing pre-weighted data to weighted data 
  + standardized tables for both matched and weighted data to compare to initial data set
