---
title: "Midterm Group Project, <br>Stats 506, F20"
author: "Hongfan Chen"
date: "`r format.Date(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    code_folding: hide
    theme: united
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, warning=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
})
```

### README

In this `SAS` tutorial, we mainly use the following procedure. `proc import` and `macros` `csvexport`(Contributed by Dr.Henderson) are used to read and export the data. `proc logistic` is used to fit the logistic regression model and predicts propensity score, and `gmatch`(Contributed by The Division of Biostatistics at the Mayo Clinic) perfroms the greedy matching.   
     
Then we use `proc gchart` to reveal the structure of the data, thereafter using `proc means` along with `proc ttest` to compute means and construct the t-test with sample weight.     
        
This work is done primarily in `SAS`, with the exception that the write up and figures or tables  for visualization are produced in R.

### SAS Script

<details>
 <summary> `GroupProject_SAS_Hongfan.sas` </summary>
```{r GroupProject_SAS_Hongfan.sas, comment=""}
readLines("./GroupProject_SAS_Hongfan.sas")
```
</details>

### Load Files

```{r load_sas, message=FALSE}
path = './'
prematch_path = sprintf('%s/output/freq_table.csv', path)
matched_path = sprintf('%s/output/freq_table_merge.csv', path)
inverseweight_path = sprintf('%s/output/ps_match_inverse_table.csv', path)
matched_sum_path = sprintf('%s/output/matched_summary.csv', path)
pre_sum_path = sprintf('%s/output/pre_summary.csv', path)
conflimits_path = sprintf('%s/output/conflimits.csv', path)
ttest_path = sprintf('%s/output/ttest.csv', path)

age_pre = sprintf('%s/graph/age_before.png', path)
age_after = sprintf('%s/graph/age_after.png', path)

hyper_pre = sprintf('%s/graph/hyper_before.png', path)
hyper_after = sprintf('%s/graph/hyper_after.png', path)

smoke_pre = sprintf('%s/graph/smoke_before.png', path)
smoke_after = sprintf('%s/graph/smoke_after.png', path)

press1_pre = sprintf('%s/graph/press_before.png', path)
press1_after = sprintf('%s/graph/press_after.png', path)

press2_pre = sprintf('%s/graph/press2_before.png', path)
press2_after = sprintf('%s/graph/press2_after.png', path)

ps_distribution = sprintf('%s/graph/ps_combined.png', path)

prematch = read_delim(prematch_path, delim = ',')
matched = read_delim(matched_path, delim = ',')
inverseweight = read_delim(inverseweight_path, delim = ',')
matched_sum = read_delim(matched_sum_path, delim = ',')
pre_sum = read_delim(pre_sum_path, delim = ',')
conflimits = read_delim(conflimits_path, delim = ',')
ttest = read_delim(ttest_path, delim = ',')
```

### 1. Distribution of Propensity Score

```{r ps_sas}
knitr::include_graphics(ps_distribution)
```

### Explanation

From the figure above, we see that there is a huge difference in the distribution of propensity score of two groups. If we compare the heart attack rate directly, a misleading result may occur. Therefore, we need to implement the Propensity Score Matching/Weighting to achieve balance or comparability of treatment groups in terms of their measured pretreatment covariates, and thereby controls for confounding bias in estimating treatment effects.

### 2. Comparison Table Between Prematch Data And Macthed Data {.tabset .tabset-fade .tabset-pills}

```{r tables_sas, message=FALSE}
tab = function(df, cap = "hello"){
  df %>%
  mutate(
    `Heart Attack` = ifelse(heart_attack == 1, "YES", "NO"),
    diabetes = ifelse(diabetes == 1, "Diabetes", "Non-Diabetes")
    ) %>%
  group_by(diabetes) %>%
  mutate(pct = COUNT / sum(COUNT)) %>%
  mutate(COUNT_PCT = sprintf('%.0f (%.2f%s)', COUNT, pct * 100, "%")) %>%
  select(`Heart Attack`, diabetes, COUNT_PCT) %>%
  pivot_wider(id_cols = `Heart Attack`,
               names_from = 'diabetes', 
               values_from = 'COUNT_PCT') %>%
  knitr::kable(format = 'html', caption = cap) %>%
  kableExtra::kable_styling("striped", full_width = TRUE)
}
```

#### Before Match

```{r table_1_sas}
cap = paste0(
"**Table 1.** Proportion of Heat Attack with Pre-matched Group"
)
tab(prematch,
    cap = cap
    )
```

#### After Match

```{r table_2_sas}
cap = paste0(
"**Table 2.** Proportion of Heat Attack with Matched Group"
)
tab(matched,
    cap = cap
    )
```

#### Using Iverse Propensity Weight

```{r table_3_sas}
cap = paste0(
"**Table 3.** Proportion of Heat Attack with Inverse Propensity Weight"
)
tab(inverseweight,
    cap = cap
    )
```


### Explanation

These tables show how `Propensity Score Matching/Weighting` eliminate or reduce the effect of pre-treatment covariates on estimating treatment effect.     
We can see that before matching/weighting, the rate of heart attack in diabetes patients is around 7 times larger than the non-diabetes patients. After matching/weighting, this rate decreases to only 2-3 times larger. This tells us the diabetes does have an effect on heart attack, but not that much.

### 3. Comparison Graph Between Prematch Data And Macthed Data {.tabset .tabset-fade .tabset-pills}

#### Age

```{r age_sas}
knitr::include_graphics(age_pre)
knitr::include_graphics(age_after)
```

#### Year of Hyper

```{r hyper_sas}
knitr::include_graphics(hyper_pre)
knitr::include_graphics(hyper_after)
```

#### Year of Smoke

```{r smoke_sas}
knitr::include_graphics(smoke_pre)
knitr::include_graphics(smoke_after)
```

#### Blood Press

```{r bloodpress_sas}
knitr::include_graphics(press1_pre)
knitr::include_graphics(press1_after)
```

#### Blood Press 2

```{r bloodpress2_sas}
knitr::include_graphics(press2_pre)
knitr::include_graphics(press2_after)
```

### Explanation

These graphs give us a direct and clear sense of how `Propensity Score Matching/Weighting` balance the pre-treatment covariates. Here I select `age`, `year of smoke`, `year of hyper`,
`blood pressure` to illustrate this fact. Before matching, the distribution of these variables differs heavily on two groups. After matching, these pre-treatment covariates seems to be unified, which helps us to mimic the situations of Randomized clinical trials(RCT).

### 4. Balance Checking For All Covariates And T Test {.tabset .tabset-fade .tabset-pills}

#### MEAN/STD

```{r Balance_sas}
n = (length(names(matched_sum)) - 1) / 3
pre_data = NULL
matched_data = NULL

for(i in 1:n){
  data0 = pre_sum %>%
    select(1, 2+3*(i-1), 3+3*(i-1), 4+3*(i-1))
  names(data0) = c("diabetes", "variable", "mean", "std")
  data_wider = data0 %>% 
    mutate(mean_std = sprintf("%.3f(%.3f)", mean, std),
           diabetes = ifelse(diabetes == "Yes",
                             "Diabetes",
                             "Non-diabetes")
    ) %>%
    select(diabetes, variable, mean_std) %>%
    pivot_wider(names_from = diabetes,
                values_from = mean_std)
  pre_data = bind_rows(pre_data, data_wider)
}

for(i in 1:n){
  data0 = matched_sum %>%
    select(1, 2+3*(i-1), 3+3*(i-1), 4+3*(i-1))
  names(data0) = c("diabetes", "variable", "mean", "std")
  data_wider = data0 %>% 
    mutate(mean_std = sprintf("%.3f(%.3f)", mean, std),
           diabetes = ifelse(diabetes == "Yes",
                             "Diabetes",
                             "Non-diabetes")
           ) %>%
    select(diabetes, variable, mean_std) %>%
    pivot_wider(names_from = diabetes,
                values_from = mean_std)
  matched_data = bind_rows(matched_data, data_wider)
}

tab = left_join(pre_data, matched_data, by = "variable")
cap_tab = paste(
  "**Table 4.** *The mean of all variables before and after matching.*",
  "Numbers in parantheses are standard deviation."
)
cn = c('Covariate',
       'Non-diabetes', 'diabetes',
       'Non-diabetes', 'Non-diabetes'
       )
tab %>%
  knitr::kable(
    format = 'html', 
    escape = FALSE, 
    align = 'lcccc',
    col.names = cn,
    cap = cap_tab
  ) %>%
  kableExtra::kable_styling("striped", full_width = TRUE) %>%
  kableExtra::add_header_above(header = c(' ' = 1,
                                          'Before Matching' = 2,
                                          'After Matching' = 2)
                               )
```

#### T-TEST

```{r ttest_sas}
tab1 = conflimits %>%
  select(Variable, Class, Method, Variances, Mean) %>%
  left_join(ttest %>% 
              select(-Variable),
            by = c("Method", "Variances")
            ) %>%
  replace_na(list(Method = "-",
                  Variances = "-",
                  tValue = "-",
                  DF = "-",
                  Probt = "-")
             )
cap_tab1 = paste(
  "**Table 5.** *Perform t test to show significance using two method.*",
  "Diabetes indeed has an effect on heart attack."
)
tab1 %>%
  knitr::kable(format = 'html', caption = cap_tab1) %>%
  kableExtra::kable_styling("striped", full_width = TRUE)
```

### Explanation

Finally, I provides two tables here. The first one is the mean-std pairs of all covariates which shows the difference of two groups. The second one shows the t test result, which indicates that after matching, the heart attack rate in diabetes patients is still higher than the non-diabetes patients.     
So now we can answer the question at the beginning: Adult patients with diabetes do have higher risk for heart attack (myocardial infarction) in the United States.