---
title: "Stats 506, F20, Midterm Project Python Write-up"
author: "Zhihao Xu, xuzhihao@umich.edu"
date: "`r format.Date(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, echo = TRUE)
library(reticulate)
library(tidyverse)
use_python("/opt/anaconda3/bin/python")
```


In this `python` tutorial, we mainly use the following package. `numpy` and 
`pandas` are used to read and process the data, `sklearn.linear_model` is used 
to fit the logistic regression model, `statsmodels.stats.weightstats` is used 
to construct the t-test with sample weight. All the algorithms related to 
propensity score weighting in this tutorial are implemented in `python`, 
but the tables and figures are visualized by `R(tidyverse)` using the 
output extracted from `python`. All the `python` variables are stored in a `R` 
list called `py` automatically and can be extracted by `$`.


```{python load_package, engine.path="/opt/anaconda3/bin/python"}
import numpy as np
import pandas as pd
from sklearn.linear_model import LogisticRegression
from statsmodels.stats.weightstats import ttest_ind
```


```{python read_data, engine.path="/opt/anaconda3/bin/python"}
nhanes = pd.read_csv('../data/nhanes.csv')
nhanes_X = nhanes.drop(columns=['id', 'heart_attack','diabete','weight'])
nhanes_diab = nhanes['diabete']
weight = nhanes['weight']
```


#### Estimate the Propensity Score by Fitting a Logistic Regression Model
```{python lg, engine.path="/opt/anaconda3/bin/python"}
lg = LogisticRegression(random_state=0, max_iter = 1000)
lg.fit(nhanes_X, nhanes_diab, sample_weight = weight)
prop_score = lg.predict_proba(nhanes_X)[:,1]
```

```{r py_psplot, fig.cap=cap,fig.height=4,fig.width=8,echo=FALSE}
cap = "**Figure 1:** Propensity score distribution by Treated/Control Groups"
ps = data.frame(
  p_score = py$prop_score,
  diabete = ifelse(py$nhanes_diab, "Diabetes", "No Diabetes")
)

ps %>%
  ggplot( aes(x = p_score) ) + 
  geom_histogram( aes(color = diabete, fill = diabete),
                      position = "identity", bins = 30, alpha = 0.3) +
  xlab("Propensity Score") + 
  ylab("Frequency") +
  theme_bw()
```

Through **Figure 1**, we can see that, compared with diabetes patient, 
the non-diabetes patients usually have lower propensity scores. 
If we compare the `heart attack` rate directly, it is very likely to 
lead to a misleading result. Hence, we need to implement the Propensity 
Score Weighting/Matching to reduce the impact of covariates.

#### Propensity Score Matching by Nearest Neighbor
```{python psmatch}
dia_idx = np.where(nhanes['diabete'].values==1)
non_dia_idx = np.where(nhanes['diabete'].values==0)
prop_score_logit = np.log(prop_score / (1 - prop_score))
std = np.std(prop_score_logit[dia_idx])
result = [0]*len(prop_score_logit[dia_idx])
for i in range(len(prop_score_logit[dia_idx])):
    dif = prop_score_logit[dia_idx][i] - prop_score_logit[non_dia_idx]
    dif[np.array(result)[np.array(result)!=0]] = 100
    min_val = min(abs(dif))
    if min_val > 0.2*std:
        result[i] = 0
    else:
        result[i] = np.where(abs(dif)==min_val)[0][0]
        
result = np.array(result)
dia_idx_matched = dia_idx[0][result!=0]
result = result[result!=0]
matched_idx = non_dia_idx[0][result]
heart_matched = nhanes['heart_attack'].values[matched_idx]
heart_non_dia = nhanes['heart_attack'].values[non_dia_idx]
heart_dia_matched = nhanes['heart_attack'].values[dia_idx_matched]
```


```{python ttest}
ttest_match = ttest_ind(heart_dia_matched, heart_matched, usevar='unequal', 
                        weights=(weight[dia_idx_matched], weight[matched_idx]))
```

#### Estimate the Propensity Weights
```{python psweight}
ps_weight = nhanes['diabete']/prop_score + (1 - nhanes['diabete'])/(1 - prop_score)
```

#### Balance Checking
```{python balcheck}
col = list(nhanes_X.columns)
result_mean_after = []
result_sd_after = []
for i in nhanes_X.columns:
    re_m = [np.average(nhanes_X[i][matched_idx],
                       weights=nhanes['weight'][matched_idx]), 
            np.average(nhanes_X[i][dia_idx_matched],
                       weights=nhanes['weight'][dia_idx_matched])]
    re_sd = [np.sqrt(np.average((nhanes_X[i][matched_idx]-re_m[0])**2, 
                                weights=nhanes['weight'][matched_idx])), 
             np.sqrt(np.average((nhanes_X[i][dia_idx_matched]-re_m[1])**2, 
                                weights=nhanes['weight'][dia_idx_matched]))]
    result_mean_after.append(re_m)
    result_sd_after.append(re_sd)

result_mean_before = []
result_sd_before = []
for i in nhanes_X.columns:
    re_m = [np.average(nhanes_X[i][non_dia_idx[0]],
                       weights=nhanes['weight'][non_dia_idx[0]]), 
            np.average(nhanes_X[i][dia_idx[0]],
                       weights=nhanes['weight'][dia_idx[0]])]
    re_sd = [np.sqrt(np.average((nhanes_X[i][non_dia_idx[0]]-re_m[0])**2, 
                                 weights=nhanes['weight'][non_dia_idx[0]])), 
             np.sqrt(np.average((nhanes_X[i][dia_idx[0]]-re_m[1])**2, 
                                 weights=nhanes['weight'][dia_idx[0]]))]
    result_mean_before.append(re_m)
    result_sd_before.append(re_sd)
```



#### {.tabset .tabset-pills .tabset-fade}
##### Table 1: Before Matching
```{r tab1, echo=FALSE}
cap1 = "**Table 1:** Proportion of Heat Attack with Original Control Group"
tab_not_match = py$nhanes %>%
  dplyr::select(heart_attack, diabete, weight) %>%
  group_by(diabete, heart_attack) %>%
  summarise(freq = sum(weight)) %>%
  mutate(
    prop = freq/sum(freq),
    freq_p = paste0(sprintf("%9.0f", freq),"(",
                    sprintf("%5.1f", prop*100),"%)")
  )%>%
  dplyr::select(-freq,-prop)

as_tibble_col(c("No","Yes"),column_name = "Heart Attack") %>%
  bind_cols(tab_not_match[1:2,3]) %>%
  bind_cols(tab_not_match[3:4,3]) %>%
  knitr::kable(format = "html", align = "lrr", caption = cap1,
               col.names = c("Heart Attack","Non-diabetes","Diabetes")) %>%
  kableExtra::kable_styling('striped', full_width = TRUE)
```

##### Table 2: After Matching
```{r tab2, echo=FALSE}
cap2 = "**Table 2:** Proportion of Heat Attack with Matched Control Group"
tab_match = py$nhanes[c(py$dia_idx_matched+1,py$matched_idx+1),] %>%
  dplyr::select(heart_attack, diabete, weight) %>%
  group_by(diabete, heart_attack) %>%
  summarise(freq = sum(weight)) %>%
  mutate(
    prop = freq/sum(freq),
    freq_p = paste0(sprintf("%9.0f", freq),"(",
                    sprintf("%5.1f", prop*100),"%)")
  )%>%
  dplyr::select(-freq,-prop)
as_tibble_col(c("No","Yes"),column_name = "Heart Attack") %>%
  bind_cols(tab_match[1:2,3]) %>%
  bind_cols(tab_match[3:4,3]) %>%
  knitr::kable(format = "html", align = "lrr", caption = cap2,
                col.names = c("Heart Attack","Non-diabetes","Diabetes")) %>%
  kableExtra::kable_styling('striped', full_width = TRUE)
```


##### Table 3: T-test
```{r tab3, echo=FALSE}
cap3 = "**Table 3:** Result of T-test"
as_tibble_row(c(tstat = py$ttest_match[[1]], 
                pvalue = py$ttest_match[[2]], 
                df = py$ttest_match[[3]] )) %>%
  knitr::kable(format = "html", align = "rr", caption = cap3,
               col.names = c("T Statistic", "p-value", "degree of freedom")) %>%
  kableExtra::kable_styling('striped', full_width = TRUE)
```



##### Table 4: Inverse Weighting
```{r tab4, echo=FALSE}
cap4 = paste0("**Table 4:** Proportion of Heat Attack with ", 
              "Inverse Propensity Score Weighting")
tab_weight = py$nhanes %>%
  dplyr::select(heart_attack, diabete, weight) %>%
  bind_cols(ps_weight = py$ps_weight) %>%
  group_by(diabete, heart_attack) %>%
  summarise(freq = sum(weight*ps_weight)) %>%
  mutate(
    prop = freq/sum(freq),
    freq_p = paste0(sprintf("%9.0f", freq),"(",
                    sprintf("%5.1f", prop*100),"%)")
  )%>%
  dplyr::select(-freq,-prop)
as_tibble_col(c("No","Yes"),column_name = "Heart Attack") %>%
  bind_cols(tab_weight[1:2,3]) %>%
  bind_cols(tab_weight[3:4,3]) %>%
  knitr::kable(format = "html", align = "lrr", caption = cap4,
                col.names = c("Heart Attack","Non-diabetes","Diabetes")) %>%
  kableExtra::kable_styling('striped', full_width = TRUE)
```
##### Table 5: Balance Checking

```{r py_balchech,echo=FALSE}
cap5 = "**Table 5:** Balance Checking Table before and after matching"
result_mean_after = do.call("rbind",py$result_mean_after)
result_sd_after = do.call("rbind",py$result_sd_after)
result_mean_before = do.call("rbind",py$result_mean_before)
result_sd_before = do.call("rbind",py$result_sd_before)
py$col %>%
  bind_cols(paste0(sprintf("%5.3f", result_mean_before[,1]),"(",
                       sprintf("%5.3f", result_sd_before[,1]),")")) %>%
  bind_cols(paste0(sprintf("%5.3f", result_mean_before[,2]),"(",
                       sprintf("%5.3f", result_sd_before[,2]),")")) %>%
  bind_cols(paste0(sprintf("%5.3f", result_mean_after[,1]),"(",
                       sprintf("%5.3f", result_sd_after[,1]),")")) %>%
  bind_cols(paste0(sprintf("%5.3f", result_mean_after[,2]),"(",
                       sprintf("%5.3f", result_sd_after[,2]),")")) %>%
  knitr::kable(format = "html", align = "lrrrr", caption = cap5,
               col.names = c("Covariate","Non-diabetes","Diabetes",
                             "Non-diabetes","Diabetes")) %>%
  kableExtra::kable_styling('striped', full_width = TRUE) %>%
  kableExtra::add_header_above(c(" " = 1, "Before Matching" = 2, 
                                 "After Matching" = 2))
```


#### Result Interpretation
Through **Table 1**, **Table 2** and **Table 4**, we can see that before 
matching/weighting, the rate of `heart attack` in diabetes patients is around 7
times larger than the non-diabetes patients. After matching/weighting, this 
rate decreases to only 2-3 times larger, which indicates that our 
matching/weighting significantly reduce the effect of the covariates. 
**Table 5** also shows similar result. Before matching, the `heart_attack` 
rate in `relative_heart_attack`, `age`, `bmi`, `blood_press`, `blood_press2`, 
`hyper_med`, `hbq_med`, `year_smoke` and `year_hyper` are significantly 
different between diabetes patients and non-diabetes patients. However, 
after matching, the rates in all corresponding covariates look similar, 
which is another evidence of the effect of propensity score matching. 
**Table 3** shows the t test result, which indicates that after matching, 
the `heart attack` rate in diabetes patients is still higher than 
the non-diabetes patients.
